\textbf{AI for science: an emerging agenda}

\emph{Working Paper from: Bridging Data-driven and Mechanistic Modelling
{[}Dagstuhl 22382{]}}

\hypertarget{summary}{%
\section{Summary}\label{summary}}

\include{summary.tex}
\hypertarget{contents}{%
\section{Contents}\label{contents}}

\protect\hyperlink{summary}{Summary \protect\hyperlink{summary}{1}}

\protect\hyperlink{_Toc121762019}{Contents
\protect\hyperlink{_Toc121762019}{3}}

\protect\hyperlink{introduction-bridging-data-driven-and-mechanistic-modelling}{Introduction:
bridging data driven and mechanistic modelling
\protect\hyperlink{introduction-bridging-data-driven-and-mechanistic-modelling}{4}}

\protect\hyperlink{snapshots-of-ai-in-science}{Snapshots of AI in
science \protect\hyperlink{snapshots-of-ai-in-science}{6}}

\protect\hyperlink{in-earth-sciences}{In Earth Sciences
\protect\hyperlink{in-earth-sciences}{6}}

\protect\hyperlink{in-environmental-and-agricultural-sciences}{In
environmental and agricultural sciences
\protect\hyperlink{in-environmental-and-agricultural-sciences}{8}}

\protect\hyperlink{in-physical-sciences}{In physical sciences 9}

\protect\hyperlink{section}{In biological sciences 11}

\protect\hyperlink{building-effective-simulations}{Building effective
simulations 15}

\protect\hyperlink{moving-upstream}{Moving upstream 15}

\protect\hyperlink{nurturing-a-diversity-of-approaches}{Nurturing a
diversity of approaches 15}

\protect\hyperlink{truth-truthiness-and-interfacing-with-the-real-world}{Truth,
truthiness, and interfacing with the real world 16}

\protect\hyperlink{connecting-simulation-to-practice}{Connecting
simulation to practice 16}

\protect\hyperlink{directions}{Directions 17}

\protect\hyperlink{connecting-data-to-causality}{Connecting data to
causality 20}

\protect\hyperlink{causality-in-science-and-data}{Causality in science
and data 20}

\protect\hyperlink{causal-models-as-a-route-to-advancing-the-science-of-ai-and-ai-for-science}{Causal
models as a route to advancing the science of AI and AI for science 20}

\protect\hyperlink{from-methods-to-application}{From methods to
application 21}

\protect\hyperlink{directions-1}{Directions 22}

\protect\hyperlink{encoding-domain-knowledge}{Encoding domain knowledge
25}

\protect\hyperlink{wheres-my-science-jetpack}{Where's My {[}Science{]}
Jetpack? 25}

\protect\hyperlink{encoding-domain-knowledge-through-model-design}{Encoding
domain knowledge through model design 25}

\protect\hyperlink{scientific-centaurs}{Scientific centaurs 26}

\protect\hyperlink{enabling-communication-across-domains}{Enabling
communication across domains 27}

\protect\hyperlink{directions-2}{Directions 28}

\protect\hyperlink{a-research-agenda-in-ai-for-science}{A research
agenda in AI for science 32}

\protect\hyperlink{accelerating-progress-in-ai-for-science}{Accelerating
progress in AI for science 34}

\protect\hyperlink{advance-new-methods-and-applications}{Advance new
methods and applications 34}

\protect\hyperlink{invest-in-tools-and-toolkits}{Invest in tools and
toolkits 35}

\protect\hyperlink{build-capability-across-disciplines}{Build capability
across disciplines 35}

\protect\hyperlink{grow-communities-of-research-and-practice}{Grow
communities of research and practice 36}

\protect\hyperlink{ai-and-science-building-the-interface}{AI and
science: building the interface 37}

\protect\hyperlink{annex-1-participants-at-machine-learning-for-science-bridging-data-driven-and-mechanistic-modelling-dagstuhl-seminar-22382-18-23-september-2022}{Annex
1: Participants at Machine Learning for Science: Bridging Data-driven
and Mechanistic Modelling (Dagstuhl Seminar 22382, 18-23 September 2022)
39}

\protect\hyperlink{_Toc121762051}{Annex 2: Research questions arising
from the `AI for science research agenda' discussion during the Dagstuhl
workshop \protect\hyperlink{_Toc121762051}{40}}

\hypertarget{introduction-bridging-data-driven-and-mechanistic-modelling}{%

\include{introduction-bridging-data-driven-and-mechanistic-modelling}

\include{snapshots-of-ai-in-science}

\include{building-effective-simulations}
\include{connecting-data-and-causality}
\include{encoding-domain-knowledge}


\hypertarget{annex-1-participants-at-machine-learning-for-science-bridging-data-driven-and-mechanistic-modelling-dagstuhl-seminar-22382-18-23-september-2022}{%
\section{Annex 1: Participants at Machine Learning for Science: Bridging
Data-driven and Mechanistic Modelling (Dagstuhl Seminar 22382, 18-23
September
2022)}\label{annex-1-participants-at-machine-learning-for-science-bridging-data-driven-and-mechanistic-modelling-dagstuhl-seminar-22382-18-23-september-2022}}

\emph{Organisers: Philipp Berens, Kyle Cranmer, Neil Lawrence, Jessica
Montgomery, Ulrike von Luxburg.}

Thank you to all those who contributed to Dagstuhl Seminar 22382,
discussions at which are the foundation for this paper:

Mauricio A Álvarez (University of Manchester, GB)

Bubacarr Bah (AIMS South Africa -- Cape Town, ZA)

Jessica Beasley (Collective Next -- Boston, US)

Philipp Berens (Universität Tübingen, DE)

Maren Büttner (Helmholtz Zentrum München \& Universität Bonn)

Kyle Cranmer (University of Wisconsin -- Madison, US)

Thomas G. Dietterich (Oregon State University -- Corvallis, US)

Carl Henrik Ek (University of Cambridge, GB)

Stuart Feldman (Schmidt Futures -- New York, US)

Asja Fischer (Ruhr-Universität Bochum, DE)

Philipp Hennig (Universität Tübingen, DE)

David W. Hogg (New York University, US)

Christian Igel (University of Copenhagen, DK)

Samuel Kaski (Aalto University, FI)

Ieva Kazlauskaite (University of Cambridge, GB)

Hans Kersting (INRIA -- Paris, FR)

Niki Kilbertus (TU München, DE \& Helmholtz AI München, DE)

Neil D. Lawrence (University of Cambridge, GB)

Gilles Louppe (University of Liège, BE)

Jakob Macke (Universität Tübingen, DE)

Dina Machuve (DevData Analytics -- A, TZ)

Eric Meissner (University of Cambridge, GB)

Siddharth Mishra-Sharma (MIT -- Cambridge, US)

Jessica Montgomery (University of Cambridge, GB)

Jonas Peters (University of Copenhagen, DK)

Aditya Ravuri (University of Cambridge, GB)

Markus Reichstein (MPI für Biogeochemistry -- Jena, DE)

Bernhard Schölkopf (MPI für Intelligente Systeme -- Tübingen, DE)

Francisco Vargas (University of Cambridge, GB)

Soledad Villar (Johns Hopkins University -- Baltimore, US)

Ulrike von Luxburg (Universität Tübingen, DE)

Verena Wolf (Universität des Saarlandes -- Saarbrücken, DE)

The Accelerate Programme for Scientific Discovery would like to thank
Schmidt Futures for its continuing support, and the donation that
enables its work.

\protect\hypertarget{_Toc121762051}{}{}\textbf{Annex 2: Research
questions arising from the `AI for science research agenda' discussion
during the Dagstuhl workshop}

Building AI systems for science

\begin{itemize}
\item
  How can AI systems accurately generalise from finite observations? How
  can they detect causality or structure from finite observations?
\item
  What is the computational cost of complexity, and what methods can
  help manage this?
\item
  What forms of system calibration and uncertainty quantification are
  useful in the context of scientific discovery? Are theoretical
  guarantees necessary?
\item
  What new forms of explainability or interpretability could facilitate
  the deployment of AI in science?
\item
  How could AI support generalisation from a small number of
  observations? What methods could enable few- or one-shot learning?
\item
  How can AI researchers build meaningful models from data to accurately
  represent causal mechanisms in the system of study? How can
  researchers identify the most effective model for their system of
  study?
\item
  What does it mean to understand a model? How can researchers combine
  explainability with complexity?
\item
  How can AI methods be made robust and easy to use in deployment by
  domain scientists?
\item
  How can advances in simulation methods be applied in domains where the
  system at hand is less easily described by equations?
\item
  What advances are needed to expand the use of simulations in science?
  How can AI help simulate laboratory experiments or environments,
  helping make more efficient different elements of the scientific
  process? How might this be expanding in the long-term, for example to
  planning experimental design or helping identify where data is
  missing?
\item
  How can `digital siblings' be used to explore the impact of different
  interventions on complex systems?
\end{itemize}

Combining human and machine intelligence

\begin{itemize}
\item
  How can AI researchers best extract, formalise and assimilate the
  knowledge that domain researchers have acquired? What forms of
  knowledge representation can formalise scientific understandings of
  the world, translating these to objective functions for AI systems?
  What forms of human-AI engagement can make use of the `qualitative'
  knowledge -- or intuitions about a system -- that domain researchers
  have accumulated?
\item
  How can AI capture the qualitative understanding that researchers have
  of their domain to more accurately or effectively characterise a
  system?
\item
  How can AI be effectively deployed to mine the existing research
  knowledge base -- for example, papers, databased, and so on -- to
  extract new insights?
\item
  Where can automation support research progress? Which elements of the
  scientific process could be automated, and where is human input vital?
\item
  What forms of collaboration are needed to effectively specify helpful
  outputs from an AI system?
\item
  How can insights from AI analysis be returned to researchers in an
  actionable way? What mix of AI design, engineering, social
  interaction, and education can make effective interfaces between
  domain researchers and AI systems?
\item
  How can the outputs of AI systems be made interpretable for scientific
  users?
\item
  How can AI researchers better understand and design for the forms of
  interpretability that resonate with domain researchers?
\item
  What processes of collaboration or co-design can help describe what
  scientists `need to know from an AI system?
\item
  What best practices or methods can be deployed to effectively
  communicate uncertainty from AI systems to human users?
\end{itemize}

Influencing practice and adoption

\begin{itemize}
\item
  What are the craft skills in AI for science? What `know how' is
  necessary to make AI work effectively in practice?
\item
  What skills-building or forms of outreach can help take AI tools out
  of the AI community and into `the lab'?
\item
  How has machine learning been used most effectively for research and
  innovation? What best practices, or lessons, do existing efforts in AI
  for science offer?
\item
  Which AI tools are suitable for which purposes, disciplines, or
  experimental designs? Is it possible to create a taxonomy for science?
\item
  Are there generalisable methods or conclusions that can be taken from
  domain-specific efforts to deploy AI for science?
\end{itemize}
